Effectful Semantics for Substructural SSA

### 2023-11-12 pldi24/paper.tex
Tap at SSA text

In Figure \ref{fig:ssa-grammar}, we present a formal grammar for our SSA syntax.
We begin by giving a grammar for \textit{types} \(A, B, C\), parametrized by a
set of \textit{base types} \(X\). We require the existence of a \textit{unit
type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products
\(A \otimes B\); we write \(\otimes\) to underscore the fact that types are
allowed to be substructural.

We separate program fragments into 3 main syntactic categories:
\textit{expressions} \(a, b, c, e\), \textit{blocks} \(s, t\), and
\textit{label-sets} \(L\). While we make a distinction between \textit{pure} and
\textit{impure} expressions, this is done at the level of typing rules rather
than via syntax. A \textit{block} is conceptualized as a series of
\textit{let-bindings} followed by a \textit{terminator}. Here, a terminator is
either an unconditional branch to another block with an argument or an
if-statement. A block may be encapsulated within a \textit{where-block}, which
defines other blocks the inner block can branch to; these blocks can also branch
to each other in a mutually recursive fashion. Similarly to blocks, expressions
may also contain let-bindings; this is somewhat of a departure from classical
SSA, in which there is only one kind of variable binding, but facilitates
various optimizations and corresponds more closely to the in-memory
representation, where pure expressions are represented as E-graph nodes. Note
that we do \textit{not} give a grammar or semantics for top-level functions,
which we can view as simply a list of arguments and a block. Except for
inlining, we will only consider the optimization and analysis of a single
function body at a time.

We proceed to give a grammar for \textit{contexts} \(\Gamma\), which consist of
a list of variables \(x\) given a type \(A\) annotated with a \textit{quantity}
\(q \subseteq \tint\). This tells us whether the variable is \textit{affine}
(i.e., can be ignored) and/or \textit{relevant} (i.e., whether it can be used
multiple times). A variable which is both affine and relevant is called
\textit{intuitionistic}. Whether we can use a variable also depends on the
\textit{linearity} of its type.

We can then give a grammar for \textit{label-contexts} \(\ms{L}\), which consist
of a list of \textit{labels} each annotated with a context \(\Gamma\), denoting
the variables which must be live on entry to the block (corresponding to the set
a dominance frontier would compute), and an \textit{argument} \(A\).

@@

We can now list out our primary typing judgements, which are given in Figure
\ref{fig:ssa-judgements}. We have three main typing judgements:
\(\hasty{\Gamma}{p}{x}{A}{q}\) is the typing judgement for expressions, which
says that the expression \(a\) has type \(A\) in context \(\Gamma\). The
annotation \(q\) means that the expression itself has quantity \(q \subseteq
\tint\), while the annotation \(p\) means that the expression has purity \(p\)
(1 for pure, 0 for impure). \(\haslb{\Gamma}{t}{\ms{L}}\) is the typing block
for blocks, which says that, using variables in \(\Gamma\), the block \(t\)
exits by jumping to labels in \(\ms{L}\). Finally,
\(\lhaslb{\ms{L}}{L}{\ms{K}}\) says that the label-set \(L\) defines the labels
\(\ms{L}\) in terms of the labels \(\ms{K}\) (blocks in \(L\) are allowed to
recursively call into \(\ms{L}\) or exit to \(\ms{K}\)).

To define these, we will need a few basic structural judgements.
\(\csplits{\Gamma}{\Delta}{\Xi}\) means that the context \(\Gamma\) can be split
into contexts \(\Delta\) and \(\Xi\). We use the notion of splitting, rather
than weakening, a context as primitive to be able to deal with linearity, since
not all variables can be dropped or duplicated. This can be used to define
weakening \(\cwk{\Gamma}{\Delta}\) as simply "splitting \(\Gamma\) into
\(\Delta\) and the empty context", since a split is allowed to drop affine
variables. Finally, \(\lwk{\ms{L}}{\ms{K}}\) means that the label-set \(\ms{L}\)
weakens to the label-set \(\ms{K}\), i.e. that \(\ms{K}\) has more labels, but
for shared labels, needs less variables to be live (but the same non-affine
variables!) The typing rules for these judgements are given in Figure
\ref{fig:ssa-structural}.

>>>>

The grammar for our syntax is given in Figure \ref{fig:ssa-grammar}, while the
typing judgements themselves are given in \ref{fig:ssa-judgements}. We begin by
giving a grammar for \textit{types} \(A, B, C\), parametrized by a set of
\textit{base types} \(X\); all syntactically well-formed types are also
semantically valid. We postulate the existence of a \textit{unit type}
\(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products \(A
\otimes B\); we write \(\otimes\) to underscore the fact that types are allowed
to be substructural, assigning every type a linearity \(\ms{lin}(A)\) satisfying
the constraints in Figure \ref{fig:ssa-structural}. In particular, the rules
\rle{unit-lin} and \rle{bool-lin} state that both the unit and boolean types
must be intuitionistic; the rule \rle{pair-lin} says that the linearity of a
pair is simply the intersection of the linearity of its components. The rules
\rle{nil-lin} and \rle{cons-lin} together say that the linearity of a context is
the intersection of the linearity of its variables, where the linearity of a
variable is the intersection of the linearity of its type and its quantity
annotation.

Our grammar proper is split into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
  is of type \(A\) with \textit{purity} \(p = 1\) if \textit{pure}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
  for this judgement are given in Figure \ref{fig:ssa-term-typing}; in
  particular, we note that a impure terms cannot be \textit{nested}; that is,
  all proper subterms of any term are pure.
  \item \textit{Blocks} \(s, t\), which are typed via the judgement
  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
  with respect to the context \(\Gamma\) and branches only to labels in
  \(\ms{L}\). The rules for this judgement are given in Figure
  \ref{fig:ssa-block-typing}.
  \item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
\end{itemize}
Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

\TODO{segue to label-contexts \(\ms{L}\)}

The rule \rle{join-nil} is the analogue to \rle{split-nil} for \textit{label-contexts}, providing a base case in that the empty label weakens into itself. The rule \rle{join-cons} says that if \(\ms{L}\) weakens into \(\ms{K}\) and \(\Delta\) weakens into \(\Gamma\), then \(\lwk{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}, \llhyp{\ell}{\Delta}{A}}\), allowing us to "forget" any extra variables dropped by \(\cwk{\Gamma}{\Delta}\). Finally, the rule \rle{join-zero} simply says that if \(\lwk{\ms{L}}{\ms{K}}\), then \(\ms{L}\) also weakens to any extension of \(\ms{K}\).

\TODO{label-weakening is backwards}

@@
